Beef Pricing Dynamics in the Canadian Grocery Market: A Comparative Analysis of Walmart and T&T

### 2024-11-28 paper/paper.qmd
charts formatted, appendix formatted

---
title: "Beef Pricing Dynamics in the Canadian Grocery Market: A Comparative Analysis of Walmart and T&T"
subtitle: "Evidence of Cultural and Strategic Influences on Beef Pricing"
author: 
  - Ziheng Zhong
thanks: "Code and data are available at: [https://github.com/iJustinn/Canadian_Grocery_Analysis.git](https://github.com/iJustinn/Canadian_Grocery_Analysis.git)."
date: today
date-format: long
abstract: "This study analyzes beef pricing dynamics in the Canadian retail market, focusing on two vendors: Walmart and T&T. Using Bayesian linear regression, we found that historical pricing and vendor-specific strategies significantly influence current beef prices, with Walmart displaying more price stability and T&T showing greater variability. These findings reveal how both market-wide and vendor-specific factors shape beef pricing, offering insights that could help retailers refine their pricing strategies and support policymakers in addressing food affordability. This paper provides a clearer understanding of the interplay between cultural preferences and retail strategies in determining food prices."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

@@

```{r}
#| label: fig-current_price
#| fig-cap: "Boxplot of the distribution of current prices"
#| echo: false
#| warning: false
#| message: false

ggplot(beef_data, aes(y = current_price)) +
  geom_boxplot(fill = "lightgrey", color = "darkgrey", alpha = 0.7) +
  labs(title = "Boxplot of the Distribution of Current Prices", y = "Current Price (in $)") +
  theme_minimal()

@@

## Model overview

To model the current price of beef, $y_i$, at time $i$, we use a Bayesian linear regression model implemented with the `stan_glm` function in the R package `rstanarm`. The response variable, $y_i$, represents the **current price of beef** in dollars. Our predictors include $x_1$, $x_2$, and $x_3$, which represent the **month**, **old price**, and **vendor**, respectively. Each component of the model is defined and justified below.

## Model set-up

@@

In this model:
- $\alpha$ represents the **intercept** term, capturing the baseline price of beef.
- $\beta_1$ corresponds to the **effect of the month** on the current price, allowing us to capture **seasonal variations** that may impact beef pricing. The month is treated as a categorical variable, acknowledging that different months can bring different demand or supply effects due to holidays, weather, or other seasonal factors.
- $\beta_2$ represents the **effect of the old price** on the current price. This allows us to account for price inertia or trends where past pricing influences current pricing.
- $\beta_3$ represents the **effect of the vendor**, also treated as a categorical variable. Different vendors may have distinct pricing strategies, and including vendor as a categorical predictor helps us capture this vendor-specific pricing variation.

@@

These priors are chosen to be **weakly informative**, allowing the data to speak for itself while still providing a reasonable range for the coefficients based on prior expectations. Specifically, a standard deviation of 2.5 reflects our expectation that most reasonable effects should fall within a plausible range without being overly restrictive.

For the residual standard deviation, $\sigma$, we assume an **Exponential(1)** prior:

@@

## Model justification

The choice of predictors is well-balanced in terms of complexity and appropriateness for this scenario. By including **month** and **vendor** as categorical variables, we acknowledge the significant impact that both seasonal changes and vendor-specific factors can have on the current price of beef. The inclusion of the **old price** reflects the idea of price inertia, where historical prices are likely to influence current pricing trends.

The model is neither **overly simplistic** nor **unnecessarily complex**: it captures important predictors without adding extraneous complexity that could lead to overfitting. **Month** and **vendor** are treated as categorical variables rather than grouping by more arbitrary categories, which maintains the expressiveness of our model while being parsimonious.

The model is implemented using `stan_glm` from the `rstanarm` package, a high-level interface to Stan for Bayesian modeling. **Stan** provides powerful sampling algorithms, making it an appropriate tool for fitting our Bayesian model efficiently, ensuring **robust convergence** and accurate posterior estimation.

### Model Assumptions and Limitations
- **Linearity**: The model assumes a linear relationship between the predictors and the response variable. This may be a limitation if the true relationship is nonlinear.
- **Normality of Residuals**: The residuals are assumed to follow a normal distribution. If this assumption does not hold, it could bias the model predictions.
- **Priors**: We used weakly informative priors to allow flexibility. However, if we had prior knowledge of specific pricing patterns, more informative priors could be used for better inference.

### Model Validation and Diagnostics
- **Posterior Predictive Checks**: We performed posterior predictive checks to verify that the model adequately captures the variation in the data. These checks involve comparing simulated data from the posterior distribution to the observed data to check for any discrepancies.
- **Convergence**: Model convergence was assessed using **trace plots** and the **R-hat statistic**, with all values being close to 1, indicating successful convergence.
- **Alternative Models**: We also considered fitting simpler linear models that did not include the vendor or seasonal components. These models showed significantly lower predictive performance, indicating the importance of including both the vendor and seasonal factors for accuracy.

@@

## Charts

@@

ggplot(beef_data, aes(x = current_price)) +
  geom_histogram(binwidth = 1, fill = "lightgrey", color = "darkgrey", alpha = 0.7) +
  geom_density(aes(y = ..count..), color = "red", size = 1) +
  labs(x = "Current Price (in $)", y = "Frequency") +
  theme_minimal()

@@

ggplot(avg_price_per_date, aes(x = date, y = avg_current_price)) +
  geom_line(color = "darkgrey", size = 1) +
  labs(x = "Date", y = "Average Current Price (in $)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

@@

ggplot(beef_data, aes(x = price_per_unit)) +
  geom_histogram(binwidth = 0.1, fill = "lightgrey", color = "darkgrey", alpha = 0.7) +
  geom_density(aes(y = after_stat(count)), color = "red", linewidth = 1) +
  labs(x = "Unit Price ($ per 100 grams)", y = "Frequency") +
  theme_minimal()

@@

## Model

@@

```{r}
#| label: tbl-raw_price_per_unit
#| tbl-cap: "Price per Unit Data for TandT and Walmart"
#| tbl-subcap: ["Raw Data", "Cleaned Data"]
#| echo: false
#| warning: false
#| message: false
#| layout-ncol: 2

@@

## Posterior predictive check
In @fig-ppcheckandposteriorvsprior-1, we implement a posterior predictive check. This plot shows the observed data distribution (in dark lines) compared with simulated datasets from the posterior (in light grey lines). The purpose of this check is to evaluate how well the model can reproduce the observed data. The close alignment between the simulated posterior predictive distributions and the observed data suggests that the model is accurately capturing the underlying patterns in the data. Specifically, we observe that the peaks and troughs in the observed data are well represented by the posterior samples, indicating that the model is capable of capturing key features such as the general shape and spread of the data distribution. The variability across the posterior predictive samples is consistent with the observed variability, suggesting no significant model misfit.

In @fig-ppcheckandposteriorvsprior-2, we compare the posterior with the prior distributions for each parameter. The posterior distribution (shown on the left) is more concentrated compared to the prior (shown on the right), which demonstrates how the data has influenced and updated our beliefs about the model parameters. For instance, the posterior distribution for the intercept has shrunk considerably compared to its prior, indicating that the data has substantially informed our estimate of the baseline price level. Similarly, for the other parameters (month, old_price, sigma, and vendorWalmart), the posteriors are less dispersed than the priors, suggesting a meaningful reduction in uncertainty. Notably, the posterior intervals are much narrower, especially for the intercept and old price parameters, indicating that the data has provided strong information about these effects. The results demonstrate that the priors were sufficiently weak, allowing the data to dominate the inference and significantly refine the parameter estimates.

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

@@

# Comparing posterior with prior
posterior_vs_prior(beef_model, color_by = "parameter") + theme_minimal() + scale_fill_grey() + scale_color_grey()
```

## Diagnostics
In @fig-stanareyouokay-1, we present a trace plot. This plot shows the sampled values for each parameter across the MCMC iterations. The trace plot helps assess whether the chains have converged to a stationary distribution. In these plots, we see that each parameter's chains are well mixed, with no discernible trends, indicating that the MCMC algorithm has adequately explored the parameter space. The lack of significant upward or downward drift suggests that all chains are consistently sampling from the target distribution, which is a strong indicator of convergence.

@@

In @fig-stanareyouokay-2, we present an Rhat plot. The Rhat statistic (also known as the potential scale reduction factor) measures the ratio of the variance between chains to the variance within chains, and values close to 1 indicate that the chains have converged. In this plot, we see that all Rhat values for the parameters are very close to 1, specifically below the threshold of 1.05, suggesting that convergence has been achieved across all parameters.

@@

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

# Rhat plot
plot(beef_model, plotfun = "rhat") + theme_minimal()
```

>>>>

---
title: "Beef Pricing Dynamics in the Canadian Grocery Market: A Comparative Analysis of Walmart and T&T"
subtitle: "Evidence of Cultural and Strategic Influences on Beef Pricing"
author: 
  - Ziheng Zhong
thanks: "Code and data supporting this analysis is available at: [https://github.com/iJustinn/Canadian_Grocery_Analysis.git](https://github.com/iJustinn/Canadian_Grocery_Analysis.git)."
date: today
date-format: long
abstract: "This study analyzes beef pricing dynamics in the Canadian retail market, focusing on two vendors: Walmart and T&T. Using Bayesian linear regression, we found that historical pricing and vendor-specific strategies significantly influence current beef prices, with Walmart displaying more price stability and T&T showing greater variability. These findings reveal how both market-wide and vendor-specific factors shape beef pricing, offering insights that could help retailers refine their pricing strategies and support policymakers in addressing food affordability. This paper provides a clearer understanding of the interplay between cultural preferences and retail strategies in determining food prices."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

The estimand of this study is the current price of beef, determined by factors such as vendor type, historical pricing, etc. By estimating how these variables impact the current pricing, the paper aims to provide a comprehensive understanding of the pricing mechanisms that affect retail beef prices in Canada.

@@

```{r}
#| label: fig-current_price
#| fig-cap: "Boxplot of the distribution of current prices"
#| echo: false
#| warning: false
#| message: false
#| fig-width: 5

ggplot(beef_data, aes(y = current_price)) +
  geom_boxplot(fill = "lightgrey", color = "darkgrey", alpha = 0.7) +
  labs(y = "Current Price (in $)") +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

@@

## Overview

To model the current price of beef, $y_i$, at time $i$, we use a Bayesian linear regression model implemented with the `stan_glm` function in the R package `rstanarm`. The response variable, $y_i$, represents the current price of beef in dollars. Our predictors include $x_1$, $x_2$, and $x_3$, which represent the month, old price, and vendor, respectively. Each component of the model is defined and justified below.

## Set-up

@@

In this model:
- $\alpha$ represents the intercept term, capturing the baseline price of beef.
- $\beta_1$ corresponds to the effect of the month on the current price, allowing us to capture seasonal variations that may impact beef pricing. The month is treated as a categorical variable, acknowledging that different months can bring different demand or supply effects due to holidays, weather, or other seasonal factors.
- $\beta_2$ represents the effect of the old price on the current price. This allows us to account for price inertia or trends where past pricing influences current pricing.
- $\beta_3$ represents the effect of the vendor, also treated as a categorical variable. Different vendors may have distinct pricing strategies, and including vendor as a categorical predictor helps us capture this vendor-specific pricing variation.

@@

These priors are chosen to be weakly informative, allowing the data to speak for itself while still providing a reasonable range for the coefficients based on prior expectations. Specifically, a standard deviation of 2.5 reflects our expectation that most reasonable effects should fall within a plausible range without being overly restrictive.

For the residual standard deviation, $\sigma$, we assume an Exponential(1) prior:

@@

## Justification

The choice of predictors is well-balanced in terms of complexity and appropriateness for this scenario. By including month and vendor as categorical variables, we acknowledge the significant impact that both seasonal changes and vendor-specific factors can have on the current price of beef. The inclusion of the old price reflects the idea of price inertia, where historical prices are likely to influence current pricing trends.

The model is neither overly simplistic nor unnecessarily complex: it captures important predictors without adding extraneous complexity that could lead to overfitting. Month and vendor are treated as categorical variables rather than grouping by more arbitrary categories, which maintains the expressiveness of our model while being parsimonious.

The model is implemented using `stan_glm` from the `rstanarm` package, a high-level interface to Stan for Bayesian modeling. Stan provides powerful sampling algorithms, making it an appropriate tool for fitting our Bayesian model efficiently, ensuring robust convergence and accurate posterior estimation.

### Assumptions and Limitations
- **Linearity**: The model assumes a linear relationship between the predictors and the response variable. This may be a limitation if the true relationship is nonlinear.
- **Normality of Residuals**: The residuals are assumed to follow a normal distribution. If this assumption does not hold, it could bias the model predictions.
- **Priors**: We used weakly informative priors to allow flexibility. However, if we had prior knowledge of specific pricing patterns, more informative priors could be used for better inference.

### Validation and Diagnostics
- **Posterior Predictive Checks**: We performed posterior predictive checks to verify that the model adequately captures the variation in the data. These checks involve comparing simulated data from the posterior distribution to the observed data to check for any discrepancies.
- **Convergence**: Model convergence was assessed using trace plots and the R-hat statistic, with all values being close to 1, indicating successful convergence.
- **Alternative Models**: We also considered fitting simpler linear models that did not include the vendor or seasonal components. These models showed significantly lower predictive performance, indicating the importance of including both the vendor and seasonal factors for accuracy.

@@

## Charting

@@

ggplot(beef_data, aes(x = current_price)) +
  geom_histogram(binwidth = 1, fill = "lightgrey", color = "darkgrey", alpha = 0.7) +
  geom_density(aes(y = ..count..), color = "red", size = 0.8) +
  labs(x = "Current Price (in $)", y = "Frequency") +
  theme_minimal()

@@

ggplot(avg_price_per_date, aes(x = date, y = avg_current_price)) +
  geom_point(color = "grey", size = 1.5, alpha = 0.7) +
  geom_smooth(method = "loess", color = "red", fill = "darkgrey", alpha = 0.3, se = TRUE) +
  labs(x = "Date", y = "Average Current Price (in $)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

@@

ggplot(beef_data, aes(x = price_per_unit)) +
  geom_density(fill = "lightgrey", color = "red", , linewidth = 0.8, alpha = 0.5, bw = 0.5) +
  labs(x = "Unit Price ($ per 100 grams)", y = "Density") +
  theme_minimal()

@@

## Modeling

@@

```{r}
#| label: tbl-price_per_unit
#| tbl-cap: "Price per Unit Data for TandT and Walmart"
#| tbl-subcap: ["Raw Data", "Cleaned Data"]
#| echo: false
#| warning: false
#| message: false
#| layout-ncol: 2

@@

## Posterior predictive check
In @fig-ppcheck, we implement a posterior predictive check. This plot shows the observed data distribution (in dark lines) compared with simulated datasets from the posterior (in light grey lines). The purpose of this check is to evaluate how well the model can reproduce the observed data. The close alignment between the simulated posterior predictive distributions and the observed data suggests that the model is accurately capturing the underlying patterns in the data. Specifically, we observe that the peaks and troughs in the observed data are well represented by the posterior samples, indicating that the model is capable of capturing key features such as the general shape and spread of the data distribution. The variability across the posterior predictive samples is consistent with the observed variability, suggesting no significant model misfit.

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheck
#| fig-cap: "Posterior prediction check result"

@@

In @fig-posteriorvsprior, we compare the posterior with the prior distributions for each parameter. The posterior distribution (shown on the left) is more concentrated compared to the prior (shown on the right), which demonstrates how the data has influenced and updated our beliefs about the model parameters. For instance, the posterior distribution for the intercept has shrunk considerably compared to its prior, indicating that the data has substantially informed our estimate of the baseline price level. Similarly, for the other parameters (month, old_price, sigma, and vendorWalmart), the posteriors are less dispersed than the priors, suggesting a meaningful reduction in uncertainty. Notably, the posterior intervals are much narrower, especially for the intercept and old price parameters, indicating that the data has provided strong information about these effects. The results demonstrate that the priors were sufficiently weak, allowing the data to dominate the inference and significantly refine the parameter estimates.

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-posteriorvsprior
#| fig-cap: "Comparing the posterior with the prior"

# Comparing posterior with prior
posterior_vs_prior(beef_model, color_by = "parameter") +
  theme_minimal() +
  scale_fill_grey() +
  scale_color_grey() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

## Diagnostics
In @fig-trace, we present a trace plot. This plot shows the sampled values for each parameter across the MCMC iterations. The trace plot helps assess whether the chains have converged to a stationary distribution. In these plots, we see that each parameter's chains are well mixed, with no discernible trends, indicating that the MCMC algorithm has adequately explored the parameter space. The lack of significant upward or downward drift suggests that all chains are consistently sampling from the target distribution, which is a strong indicator of convergence.

@@

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-trace
#| fig-cap: "Trace plot"

In @fig-rhat, we present an Rhat plot. The Rhat statistic (also known as the potential scale reduction factor) measures the ratio of the variance between chains to the variance within chains, and values close to 1 indicate that the chains have converged. In this plot, we see that all Rhat values for the parameters are very close to 1, specifically below the threshold of 1.05, suggesting that convergence has been achieved across all parameters.

@@

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-rhat
#| fig-cap: "Rhat plot"

# Rhat plot
plot(beef_model, plotfun = "rhat") + theme_minimal()
